Supporting Dynamic Data Partitioning in Parallel Programs

### 2012-10-31 doc/types-pldi2013/intro.tex
intro, abstract

In the past decade machine architectures, particularly at the high
performance end of the spectrum, have undergone a revolution.  The
latest supercomputers are now composed of heterogeneous processors
and deep memory hierarchies.  Programming this class of machines is
increasingly difficult.  Current programming systems for these
machines have intricate features for describing parallelism, but contain
few abstractions for describing properties of program data
(e.g. independence).  Consequently, the burden of managing
the correctness of parallel computations and movement of data
through the memory hierarchy is placed on the programmer.

@@

To combat this problem, researchers have previously explored programming
systems that are capable of reasoning about the structure of data in
parallel programs.  Traditionally, these attempts have taken one of two approaches.   
The first approach relies on fully static analyses that
have no runtime overhead, but limited expressivity.  Two recent examples, 
Sequoia \cite{Fatahalian06} and 
Deterministic Parallel Java (DPJ) \cite{Bocchino09}, each provide a 
mechanism to statically partition the heap recursively into a tree of collections 
of data.  The two designs are different in many aspects, but agree that there is
a single tree-shaped partitioning of data that must be fully checked statically 
(see Section~\ref{sec:related} for more discussion of related work).  The
second approach relies on a totally dynamic analysis of program data that doesn't
limit expressivity, but incurs runtime overhead.  Examples of this approach 
include transactional memory \cite{Harris05} and thread-level speculation \cite{Steffan00}.

These two approaches illustrate different mechanisms for dealing with
the problem of data aliasing in parallel programs.  The fully static approach disallows aliasing
by requiring a single disjoint partitioning of data.  Our own experience with writing high-performance 
applications in both Sequoia and the current industry standard mix of MPI, shared-memory
threads, and CUDA, has taught us that for many practical situations a
single, static partitioning is insufficient.  In practice, the
best way to partition data is often a function of the data
itself---i.e., the partitions need to be dynamically computed.
Furthermore, it is useful to allow multiple, simultaneous partitions of
the same data, providing different views on to that data.  Alternatively, 
the fully dynamic approach permits arbitrary data aliasing, but requires 
centralized control logic to perform conflict detection between every pair 
of parallel computations. This results in runtime overhead and is 
not scalable on distributed-memory machines.

>>>>

Recently machine architecture, particularly at the high
performance end of the spectrum, has undergone a revolution.  The
latest supercomputers are now composed of heterogeneous processors
and deep memory hierarchies.  Current programming systems for these
machines have elaborate features for describing parallelism, but 
few abstractions for describing the organization of data.  However, as the architecture
of distributed memory has become increasingly complex, the placement and movement of dat has
become a first order programming concern.

%Consequently, the burden of managing
%the correctness of parallel computations and movement of data
%through the memory hierarchy is placed on the programmer.

@@

Previous work has explored two approaches to expressing the organization of data
in distributed memory parallel programs.
One class of work uses fully static analysis 
with no runtime overhead but limited expressivity.  Two recent examples, 
Sequoia \cite{Fatahalian06} and 
Deterministic Parallel Java (DPJ) \cite{Bocchino09}, each provide a 
mechanism to statically partition the heap into a tree of collections 
of data.  The two designs are different in many aspects, but agree that there is
a single tree-shaped partitioning of data that must be fully checked statically 
(see Section~\ref{sec:related}).  The
second approach relies on fully dynamic analysis of program data that doesn't
limit expressivity but incurs runtime overhead;  examples 
include transactional memory \cite{Harris05} and thread-level speculation \cite{Steffan00}.

Our own experience writing high-performance 
applications in both Sequoia and the current industry standard mix of MPI, shared-memory
threads, and CUDA, has taught us that a
single, static data partitioning is insufficient.  In practice, the
best way to partition data is often a function of the data
itself---i.e., the partitions must be dynamically computed.
Furthermore, applications often need multiple, simultaneous partitions of
the same data, providing different views on to that data.  Alternatively, 
the fully dynamic approach permits arbitrary partitioning, but requires 
centralized control logic to perform conflict detection between every pair 
of parallel computations, resulting in runtime overhead that isn
not scalable on distributed-memory machines.

### 2012-10-31 doc/types-pldi2013/legion-types.tex
intro, abstract

\title{A Type System for Structured Aliasing in Parallel Programs}
%\title{Scalable and Safe Parallelism using Logical Regions}
\authorinfo{Sean Treichler \and Michael Bauer \and Alex Aiken}{Stanford University}{\{sjt,mebauer,aiken\}@cs.stanford.edu}
%\authorinfo{Sean Treichler}{Stanford Unviersity}{sjt@cs.stanford.edu}
%\authorinfo{Michael Bauer}{Stanford University}{mebauer@cs.stanford.edu}
%\authorinfo{Alex Aiken}{Stanford University}{aiken@cs.stanford.edu}
\maketitle

\begin{abstract}
Aliasing introduces a significant impediment to reasoning about the
structure and usage of data in parallel programs.  Traditional approaches
have either disallowed aliasing to enable static analyses, or have permitted
aliasing at the cost of dynamic runtime analyses.  In this paper we introduce a
type system for the Legion programming model that uses a hybrid static and
dynamic approach to manage aliasing with minimal runtime overhead.  We prove 
that our type system can soundly reason about the privilege and coherence 
properties in Legion programs. Furthermore, we prove that type soundness 
enables a hierarchical, distributed scheduling algorithm for Legion programs 
even in the presence of aliasing. We show that an implementation of a Legion 
type checker improves the performance of Legion programs up to 71\% by 
eliding memory checks that the type system proves safe.  We also demonstrate 
that an implementation of the Legion runtime based on the hierarchical 
scheduling algorithm scales well for three real-world applications, achieving 
up to 62.5X speedup on 96 GPUs on the Keeneland supercomputer.
%Legion is a programming model that introduces
%{\em logical regions} as a mechanism for targeting large heterogeneous machines 
%with complex memory hierarchies.  Logical regions are first-class 
%values in Legion, and programmers may {\em partition} regions in multiple different
%ways into subregions.  We present
%a type system with privileges and coherence properties for reasoning about logical regions.  
%We prove the soundness of the type system and use it to show
%that all memory accesses in Legion are safe.  Furthermore, we prove that
%type soundness enables a scalable, hierarchical
%scheduling algorithm for Legion programs.  We show that an implementation of the type checker
%for Legion programs improves performance up to 71\% by eliding
%memory checks the type system proves safe.  We also demonstrate that an implementation of the Legion 
%runtime based on the hierarchical scheduling algorithm scales well for
%three real-world applications.

>>>>

%\title{A Type System for Structured Aliasing in Parallel Programs}
\title{Supporting Dynamic Data Partitioning in Parallel Programs}
%\title{Scalable and Safe Parallelism using Logical Regions}
\authorinfo{Sean Treichler \and Michael Bauer \and Alex Aiken}{Stanford University}{\{sjt,mebauer,aiken\}@cs.stanford.edu}
%\authorinfo{Sean Treichler}{Stanford Unviersity}{sjt@cs.stanford.edu}
%\authorinfo{Michael Bauer}{Stanford University}{mebauer@cs.stanford.edu}
%\authorinfo{Alex Aiken}{Stanford University}{aiken@cs.stanford.edu}
\maketitle

\begin{abstract}
Many distributed-memeory parallel applications dynamically partition 
data, but current language designs either 
require fully static partitioning or rely on entirely dynamic techniques.
We describe Legion, a programming system with a combination
of language features and static and dynamic checks to support 
dynamic data partitions with minimal overhead.
We describe the novel aspects of the Legion design, prove the soundness of the Legion type system,
 and show Legion type checking improves performance by up to 71\% by eliding provably
safe memory checks.  Furthermore, we show how the language design and type safety
enables a hierarchical, distributed runtime scheduling algorithm, and how
a system of data coherence annotations improves expressivity without compromising
soundness. We report results for three real-world applications running
on distributed memory machines, achieving up to 62.5X speedup on 96 GPUs 
on the Keeneland supercomputer.
%Legion is a programming model that introduces
%{\em logical regions} as a mechanism for targeting large heterogeneous machines 
%with complex memory hierarchies.  Logical regions are first-class 
%values in Legion, and programmers may {\em partition} regions in multiple different
%ways into subregions.  We present
%a type system with privileges and coherence properties for reasoning about logical regions.  
%We prove the soundness of the type system and use it to show
%that all memory accesses in Legion are safe.  Furthermore, we prove that
%type soundness enables a scalable, hierarchical
%scheduling algorithm for Legion programs.  We show that an implementation of the type checker
%for Legion programs improves performance up to 71\% by eliding
%memory checks the type system proves safe.  We also demonstrate that an implementation of the Legion 
%runtime based on the hierarchical scheduling algorithm scales well for
%three real-world applications.
